---
title: "Estimating the effective reproduction number from wastewater (Rt)"
output: github_document
---

## Introduction

This repository is a collection of methods for estimating the effective reprodouction number from wastewater (the Rt). The Rt is a number that indicates the growth or decline of disease spread. Eight distinct methods were identified from the literature and subsequently compared to one another for accuracy, computation time, and efficiency. Each method has been wrapped in a function with data prep provided. On this page, you will find some descriptions of the methods and examples. Full explanations and documentation of each Rt method are available in the **functions** folder.

## Citation information

When using code or other content from this repository, please cite the following:

Dustin T. Hill,  Yifan Zhu, Christopher Dunham, Joe Moran, Yiquan Zhou, Mary Collins, Brittany Kmush, and David Larsen. Translating wastewater surveillance data into public health terms: Methods for estimating the effective reproduction number (Rt). **add preprint link**


## EpiEstim and Rt functions

Much of the work that has gone into the calculation of the Rt originates with models and methods proposed by Cori et al (2013) and their `EpiEstim` R package. 

```{r EpiEstim example, include=TRUE, echo=TRUE, message=FALSE, error=FALSE, warning=FALSE}

library(dplyr)
library(tidyr)
library(tidyverse)
library(EpiEstim)
library(ggplot2)
library(zoo)
library(scales)
library(broom)

# 
# DATA
dat <- readRDS("data/Rt_data_state.rds")

# Reference model - case based Rt
source("functions/Rt-EpiEstim-functions.R")

# example county df
cases_df <- dat %>%
  ungroup() %>%
  select(date, state_new_cases_7avg) %>%
  rename(Date = date) 

# Rt from  case data with SI 4 (not much difference in the unknown si results)
rt_cases_weekly <- rt_function(cases_df, mean_si = 4, std_si = 1, weekly = "Yes")

# start and end dates for time series
start_date <- as.Date("2022-09-01")
end_date <- as.Date("2024-02-20")

# plot
rt_case_plot <- 
  ggplot(data = rt_cases_weekly, aes(x = date, y = mean_rt))+
  geom_ribbon(aes(ymin = ll_95_rt, ymax = ul_95_rt), alpha = 0.4, fill = "orange") +
  geom_line(color = "orange", linewidth = 1) +
  geom_hline(yintercept = 1, lty = "dashed")+
  geom_vline(xintercept = as.Date("2022-12-21"), lty = "longdash")+
  geom_vline(xintercept = as.Date("2023-09-21"), lty = "longdash")+
  geom_vline(xintercept = as.Date("2024-01-05"), lty = "longdash")+
  theme_bw() +
  labs(title = expression("Statewide effective reproduction number (R"["t"]~") from case data"),
       x = "",
       y = expression("Mean R"["t"]~"cases"))+
  scale_x_date(labels = date_format("%b %y"),
               date_breaks = "1 month", limits = c(start_date, end_date))+
  theme(axis.text.x = element_text(angle = 90))
rt_case_plot



```

## Example wastewater-based Rt method

Eight wastewater methods were identified and are explained in detail in the manuscript. Examples for running each method can be found in the `functions` folder.

```{r ww example,include=TRUE, echo=TRUE, message=FALSE, error=FALSE, warning=FALSE}

# Method 1 - fit line
source("functions/Rt-fit-line-function.R")

#extract ww data and join it to rt data
ww_data <- dat %>%
  select(date, state_sars2)%>%
  arrange(date) 

# one dataframe for the county with ww data and rt from case data
data_fit_line <- inner_join(ww_data, rt_cases_weekly, c("date")) %>%
  rename(rt = mean_rt)

# run function
rt_fit_weekly <- rt_fit_line_function(dataframe = data_fit_line, weekly = "Yes", range = 45, predictor = "state_sars2")
rt_cases_weekly$model <- "Rt cases"
rt_fit_weekly$model <- "Fit line"
values <- c("orange", "darkblue")

# plot
plot_fit_line <- 
  ggplot()+
  geom_line(data = rt_cases_weekly, aes(x = date, y = mean_rt, color = model), linewidth = 1) +
  geom_ribbon(data = rt_cases_weekly, aes(x = date, y = mean_rt, ymin = ll_95_rt, ymax = ul_95_rt, fill = model), alpha = 0.4) +
  geom_line(data = rt_fit_weekly, aes(x= week, y = mean_rt, color = model), linewidth = 1)+
  geom_ribbon(data = rt_fit_weekly, aes(x = week, ymin = ll_95_rt, ymax = ul_95_rt, fill = model), alpha = 0.2) +
  geom_hline(yintercept = 1, lty = "dashed") +
  labs(title = expression("R"["t"]~"fit line"),
       x = "Date",
       y = "Rt")+
  scale_colour_manual(name = "",
                      values =values)+
  scale_fill_manual(name = "", values = values)+
  theme_bw()+
  guides(color = guide_legend(override.aes = list(linewidth= 2)))+
  scale_x_date(labels = date_format("%b %y"),
               date_breaks = "1 month", 
               #limits = c(as.Date("2023-10-01"), as.Date("2024-02-20"))
               limits = c(start_date, end_date)
               )+
  theme(axis.text.x = element_text(angle = 90),
        legend.position = "bottom")
plot_fit_line

```

## Repository outline

### Data folder

This folder contains data for generating the figures and tables published in the paper. Data are from New York State's Wastewater Surveillance Network [Open Data website](https://coronavirus.health.ny.gov/covid-19-data-new-york).

### Functions folder

This folder contains a function for each Rt method as well as additional functions used in the calculation of each Rt method. The functions are wrappers for the methods and packages that we review in the paper.

### Figure and table scripts folder

This folder contains the R scripts for generating the figures and tables included in the Hill et al. publication linked to this repository.

### High-throughput folder

This folder contains example scripts for setting up a parallel process on separate workers with HT Condor and Conda. Consult your IT department for assistance. These scripts are intended as an example and will not work without adaptation to your specific computing environment.

#### Note on high-throughput computing

Six of the methods that we evaluated can be run on most machines, however, two methods (`EpiSewer` and the Goldstein et al. method) require additional computing resources. We used a high-throughput computing approach through our institution's computing cluster that was based in HT Condor and used Conda to set up R and other environments. We provide an example of that process in the **high-throughput** folder. Depending on your institution, you may need to set up your own process. This is only needed if you intend to iterate Rt estimates over multiple separate sets of data. As an example, New York has 62 counties and we wanted to compare the accuracy of the models across counties based on population size. Thus, we needed 62 separate Rt estimates for each method. A `for` loop and a function for the six other methods, however, for `EpiSewer` and the Goldstein method, we needed to distribute each county calculation as separate processes. Thus, the need for high-throughput computing to set up 62 separate calculations. 

If you are estimating the Rt for one jurisdiction, you will likely not need high-througput computing. Note that each method varies in the length of time it takes to run, and speed will be impacted by the data structure. Please consult the individual package documentation and papers associated with each method for more details and to troubleshoot any issues.


## References and resources


### EpiEstim

Cori A, Ferguson NM, Fraser C, Cauchemez S. A New Framework and Software to Estimate Time-Varying Reproduction Numbers During Epidemics. Am J Epidemiol. 2013 Nov 1;178(9):1505–12. 

Cori A. EpiEstim: Estimate Time Varying Reproduction Numbers from Epidemic Curves [Internet]. 2021. Available from: [https://CRAN.R-project.org/package=EpiEstim](https://CRAN.R-project.org/package=EpiEstim)

GitHub page: [https://github.com/mrc-ide/EpiEstim](https://github.com/mrc-ide/EpiEstim)

### Huisman et al. method

Huisman JS, Scire J, Caduff L, Fernandez -Cassi Xavier, Ganesanandamoorthy P, Kull A, et al. Wastewater-Based Estimation of the Effective Reproductive Number of SARS-CoV-2. Environ Health Perspect. 2022;130(5):057011. 

GitHub page: [https://github.com/JSHuisman/wastewaterRe](https://github.com/JSHuisman/wastewaterRe)

### Goldstein et al. method

Goldstein IH, Parker DM, Jiang S, Minin VM. Semiparametric inference of effective reproduction number dynamics from wastewater pathogen surveillance data. ArXiv. 2023 Aug 31;arXiv:2308.15770v2. 

Isaac H Goldstein, Daniel M Parker, Sunny Jiang, Volodymyr M Minin, Semiparametric inference of effective reproduction number dynamics from wastewater pathogen surveillance data, Biometrics, Volume 80, Issue 3, September 2024, ujae074, [https://doi.org/10.1093/biomtc/ujae074](https://doi.org/10.1093/biomtc/ujae074)

GitHub page: [https://github.com/igoldsteinh/concRt](https://github.com/igoldsteinh/concRt)

### EpiSewer method

GitHub page: [https://github.com/adrian-lison/EpiSewer](https://github.com/adrian-lison/EpiSewer)

### ERN method

Champredon D, Papst I, Yusuf W. ern: An R package to estimate the effective reproduction number using clinical and wastewater surveillance data. PLOS ONE. 2024 Jun 21;19(6):e0305550. 

GitHub page: [https://github.com/phac-nml-phrsd/ern](https://github.com/phac-nml-phrsd/ern)
